---
title: "Rolleston Glacier Fieldwork map"
output: html_notebook
---

This notebook has been prepared to enable preparation of a map of where observations have been taken on the Rolleston Glacier.
Loadlibraries
```{r}
if (!require(sp)) install.packages('sp'); library(sp)
if (!require(terra)) install.packages('terra'); library(terra)
if (!require(slippymath)) install.packages('slippymath'); library(slippymath) #needed to figure out wmts tiles
if (!require(sf)) install.packages('sf'); library(sf) #needed to work with slippymath
if (!require(tools)) install.packages('tools'); library(tools) #has the very useful file_ext() function
```


Start by describing the various data sources:
Set the observation date of interest
Observation data
Map data
Spatial data
```{r}
#****************************************
#Edit the date to match the survey date
#***************************************
DateOfInterest <- as.Date("2024-11-18")

LINZAPIToken <- "85df745fa5d446fea241dd5ae40add85"
LINZTopo50WMSURL <- "https://data.linz.govt.nz/services;key=YOUR_API_TOKEN/wms?service=WMS&version=1.1.1&request=GetMap&layers=layer-50767&format=image/png&width=364&height=400&bbox=LonMin,LatMin,LonMax,LatMax"
ProjectDirectory <- "/media/drizzle/70DAC0494B655316/Projects/Rolleston Glacier/RollMassBal"

#Check directory exists, if it doesn't try the windows path.
if (!dir.exists(ProjectDirectory)){ProjectDirectory <- "D:\\Projects\\Rolleston Glacier\\RollMassBal"}
DataDirectory <- file.path(ProjectDirectory,"data")
GISDirectory <- file.path(ProjectDirectory,"GIS")

#Figure out the directory that the pit and probe data are in as the directory naming convention is based on the mass balance year range.
YearOfInterest <- as.numeric(format(DateOfInterest,"%Y"))
MonthOfInterest <- as.numeric(format(DateOfInterest,"%m"))
if (MonthOfInterest > 10){
  PitAndProbeDataDirectory <- paste0("IndividualYears/",YearOfInterest,"-",YearOfInterest + 1)
} else {
  PitAndProbeDataDirectory <- paste0("IndividualYears/",YearOfInterest - 1,"-",YearOfInterest)
}


RollestonGlacierOutlineFile <- "Rolleston_outline_20220408_Sentinel2_NDSI.shp"
ProbingFileName     <- file.path(DataDirectory,
                                 PitAndProbeDataDirectory,
                                 paste0("Rolleston_depths_",format(DateOfInterest,"%Y%m%d"),".csv"))
SnowPitFileName     <-  file.path(DataDirectory,
                                  PitAndProbeDataDirectory,
                                  paste0("Rolleston_density_",format(DateOfInterest,"%Y%m%d"),".csv"))
StakeFileName       <- file.path(DataDirectory,"Rolleston_Stakes.csv")

#If an EOSS snowline spatial file exists, then use it. Can be either an ESRI shapefile or a GPX file which the MapToaster iphone app can produce.
#EOSSFileName        <- "track-2021-03-23_09-01-55.gpx"
#EOSSFileName        <- "EOSS_20240320.shp"
```

Now lets load the Glacier outline, stake points, pit location, probe locations and plot them on the Topo50 map.
Note that the Topo50 data arrives in lat lon format, and I haven't figured out how to change that. This means I have to reproject everything to put it on the plot.
```{r,fig.height=5,fig.width=7.5}
#Load the glacier outline and explicitly set the projection to NZTM
RollestonGlacierOutline <- terra::vect(file.path(GISDirectory,RollestonGlacierOutlineFile)) %>% terra::project(y="epsg:2193")

#Get the data and convert to spatial objects. Note that stake data is for all years, so is in a different directory
if (file.exists(SnowPitFileName)) {
  PitData <- read.csv(SnowPitFileName)
  spPitPoints <- terra::vect(PitData, geom=c("Easting.NZTM.","Northing.NZTM."), crs="epsg:2193")
  } else spPitPoints <- NULL
if (file.exists(ProbingFileName)) {
  ProbeData <- read.csv(ProbingFileName)
  spProbePoints <- terra::vect(ProbeData, geom=c("Easting_NZTM","Northing_NZTM"), crs="epsg:2193")

  } else spProbePoints <- NULL
StakeData <- read.csv(StakeFileName,stringsAsFactors = FALSE,
                      colClasses = c("character","character",rep("numeric",5),"character"))
#Limit the stake data to just the date(s) of interest
StakeData <- StakeData[as.Date(StakeData$Date,"%d/%m/%Y") == DateOfInterest,]
spStakePoints <- terra::vect(StakeData, geom=c("E_NZTM","N_NZTM"), crs="epsg:2193")

if(exists("EOSSFileName")){
  #if it is a GPX file then process it one way
  if (tolower(file_ext(EOSSFileName)) == "gpx"){
    #Get the snow line track data. This is in GPX format. Projected in WGS84
    EOSSTrackPoints <- readGPX(file.path(GISDirectory,basename(PitAndProbeDataDirectory),EOSSFileName))[["tracks"]][[1]][[1]]
    #Convert to a spatial file
    EOSSLine <- terra:vect(as.matrix(EOSSTrackPoints[,c("lon","lat")]), type="lines",crs="epsg:4326")
  } else if (tolower(file_ext(EOSSFileName)) == "shp"){ #if its an ESRI shapefile, process accordingly
    EOSSLine <- terra::vect(file.path(GISDirectory,basename(PitAndProbeDataDirectory),EOSSFileName)) 
    }

} else {EOSSLine <- NULL}


#Put them in a list for simplicity of later processing
ObservedData <- list(PitPoints = spPitPoints, ProbePoints = spProbePoints, StakePoints = spStakePoints, Outline = RollestonGlacierOutline, EOSS = EOSSLine)

#Get the extents of the glacier in lat lon
GlacierExtentLatLon <- terra::project(RollestonGlacierOutline, crs('+init=epsg:4326')) %>% terra::ext()
#Expand the extents to provide extents for the topographical map to download
MapExtents <- GlacierExtentLatLon + 0.005

#And use those extents to get the Topo50 data. Note that I seem to need to download the data first, then load it into R.
#LINZ now provides TOPO50 data via a WMTS (not WMS) so the tiles that are wanted needed to be figured out before they are requested. 
#It seems there is a standard projection (world mercator, EPSG:3857) used for WMTS

WMTS_URL <- "https://tiles-cdn.koordinates.com/services;key=8ed417cc81ea45a0b92d597307229b80/tiles/v4/layer=50767/EPSG:3857/{z}/{x}/{y}.png"

#What follows is derived from the slippymaths package
#Create a vector of the bounding box
sf_bbox <- sf::st_bbox(MapExtents)

#Figure out the tile numbers and zoom level
tile_grids <- bbox_to_tile_grid(sf_bbox, max_tiles = 4)

#Figure out the coordinates of the extents of each of the tiles
tile_bboxes <- tile_grid_bboxes(tile_grids)

#Download each tile image
Tile_images <- lapply(seq(1,nrow(tile_grids$tiles)), function(tile){
  
  Tile_URL <- sub("\\{z\\}",tile_grids$zoom,WMTS_URL)
  Tile_URL <- sub("\\{x\\}",tile_grids$tiles[tile,'x'],Tile_URL)
  Tile_URL <- sub("\\{y\\}",tile_grids$tiles[tile,'y'],Tile_URL)

  Image <- terra::rast(Tile_URL)
  
  #Set the real extent of the tile
  #raster::extent(Image) <- raster::extent(tile_bboxes[[tile]][c("xmin", "xmax", "ymin", "ymax")])
   terra::ext(Image) <- terra::ext(tile_bboxes[[tile]][c("xmin", "xmax", "ymin", "ymax")])
  
  #Set the real projection
  terra::crs(Image) <- 'epsg:3857'
  
  #Convert to an RGB brick so that colors are retained when merging them later
  Image <- terra::colorize(Image, "rgb") %>% terra::flip()
  return(Image)
})
Topo50Map <- do.call(terra::merge,Tile_images)

#Reproject the other spatial data to the maps projection ready for plotting
reprojected.data <- lapply(ObservedData, function(SpatialElement) {
  if(!is.null(SpatialElement)){terra::project(SpatialElement,terra::crs(Topo50Map))
    }else {NULL}
  })
list2env(reprojected.data,env=.GlobalEnv)

Topo50Map <- terra::crop(Topo50Map, y = terra::ext(Outline) + 300)
#Reproject to the Topo50 coordinate system
#par(xpd=TRUE)
{
plotRGB(Topo50Map,colNA="white")
plot(Outline, add=TRUE)
points(ProbePoints,pch=20)
points(StakePoints, pch = 8, col = "dark green")
points(PitPoints, pch = 0, cex=1.3, col = "brown")
lines(EOSS, col = "black",lty=2)
if(!is.null(EOSSLine)){
legend("top",bty="n",legend=c("Pit","Stakes","Probes","EOSS","Glacier \nOutline"),pch=c(0,8,20,NA,NA),pt.cex=c(1.3,1,1,NA,NA),merge=TRUE,lty=c(-1,-1,-1,2,1),xpd=TRUE)
} else {
legend("topleft",inset = c(0,0),bty="n",legend=c("Pit","Stakes","Probes","Glacier \nOutline"),pch=c(0,8,20,NA),pt.cex=c(1.3,1,1,NA),merge=TRUE,lty=c(-1,-1,-1,1),col=c("brown","dark green","black","black"),xpd=TRUE)
}

terra::north(type=2,xy=c(par("usr")[1] + 100,par("usr")[3]+400),cex=1.5)
terra::sbar(d=150,xy=c(par("usr")[1] + 50,par("usr")[3]+100),divs=1,below="metres",adj=c(0.5,-1.5),labels=c("0","","150"),ticks=TRUE,xpd=TRUE)
}

#Generate a plot file
dev.copy(png,file=file.path(DataDirectory,PitAndProbeDataDirectory,
                            paste0("ObservationMap_",format(DateOfInterest,"%Y%m%d"),".png"))
         ,width=15,height=10,units="cm",res=600,family="Arial",pointsize=6)
invisible(dev.off())
```


```{r}
#if (!require(shiny)) install.packages("shiny"); library(shiny)
if (!require(leaflet)) install.packages("leaflet"); library(leaflet)
if (!require(htmltools)) install.packages("htmltools"); library(htmltools)
if (!require(raster)) install.packages("raster"); library(raster)
if (!require(rgdal)) install.packages("rgdal"); library(rgdal)
#if (!require(grDevices)) install.packages("grDevices"); library(grDevices)

NZTopo250 = 'http://tiles-a.data-cdn.linz.govt.nz/services;key=8ed417cc81ea45a0b92d597307229b80/tiles/v4/layer=52324/EPSG:3857/{z}/{x}/{y}.png'
NZTopo50 = 'http://tiles-a.data-cdn.linz.govt.nz/services;key=8ed417cc81ea45a0b92d597307229b80/tiles/v4/layer=52343/EPSG:3857/{z}/{x}/{y}.png'

#Setup the map
map <- leaflet::leaflet(options = leafletOptions(minZoom = 14)) %>% 
  setView(lng=171.525,lat=-42.89,zoom=9) %>%
  #setMaxBounds(lng1 = 171.5213, lat1 = -42.88425, lng2=171.532, lat2=-42.89375) %>% 
  addTiles(urlTemplate = NZTopo250, options = providerTileOptions(minZoom=2),attribution = "<a href=\"http://https://www.linz.govt.nz/\">LINZ</a>")

#Add the Glacier outline
map <- map %>% addPolygons()

#Add the probe locations
map <- map %>% addMarkers()

#Add the pit location
map <- map %>% addMarkers()

#Add the stake locations
map <- map %>% addMarkers()


  
```

